Computer Assisted Verbal Autopsy: Comparing Large Language Models to Physicians for Assigning Causes to 6939 Deaths in Sierra Leone from 2019-2022

### 2025-06-25 manuscript/manuscript.tex
Working on docx

VA involves two major components: survey and COD assignment \citep{worldhealthorganizationVerbalAutopsyStandards2023,chandramohanEstimatingCausesDeath2021,worldhealthorganizationVerbalAutopsyStandards2007}. In the survey component, trained interviewers use structured questionnaires and open narrative prompts to gather data from relatives or close contacts of the deceased. In the COD assignment component, physicians review these data to determine the most likely COD. However, reliance on physician assignment has been criticized for limited reproducibility and subjectivity \citep{gomesNationwideMortalityStudies2017,jhaProspectiveStudyOne2005,mccormickProbabilisticCauseofdeathAssignment2016b,morrisFactorsAssociatedPhysician2010,solemanVerbalAutopsyCurrent2006}. To overcome these limitations, automated Computer Coded Verbal Autopsy (CCVA) methods such as InterVA \citep{byassIntegratedApproachProcessing2019} and InSilicoVA \citep{mccormickProbabilisticCauseofDeathAssignment2016b} have been developed. These models offer scalable and reproducible alternatives and have demonstrated comparable performance to physicians at the population level. However, their performance at the individual level remains limited \citep{jhaAutomatedPhysicianAssignment2019,leitaoComparisonPhysiciancertifiedVerbal2014,desaiPerformanceFourComputercoded2014,tungaVerbalAutopsyModels2021,otiVerbalAutopsyInterpretation2010}, while their reliance on structured questionnaire data often omits open narrative text, which can contain additional contextual and chronological information that may improve diagnostic accuracy \citep{jebleeAutomaticallyDeterminingCause2019,blancoExtractingCauseDeath2021,kingQualityDiagnosticValue2016}.

@@

Four computational models were used to assign causes of death (CODs) for each of the 6,939 physician-agreed verbal autopsy (VA) records: GPT-3.5, GPT-4, InterVA-5, and InSilicoVA. InterVA-5 and InSilicoVA are widely used statistical models within the OpenVA framework for COD assignment in VAs \citep{murrayUsingVerbalAutopsy2014, otiVerbalAutopsyInterpretation2010, benaraEvaluationMethodsAssigning2023, chandramohanEstimatingCausesDeath2021, tungaVerbalAutopsyModels2021, leitaoComparisonPhysiciancertifiedVerbal2014, jhaAutomatedPhysicianAssignment2019, liOpenVAToolkitVerbal2023}. InterVA-5 applies a Bayesian probabilistic approach, using a standardized set of symptoms and expert-derived conditional probabilities to assign the most likely COD based on maximum probability \citep{byassIntegratedApproachProcessing2019, byassStrengtheningStandardisedInterpretation2012, bayesEssaySolvingProblem1958}. InSilicoVA extends this approach by incorporating a hierarchical Bayesian framework and Markov Chain Monte Carlo (MCMC) methods \citep{brooksMarkovChainMonte1998, chibMarkovChainMonte2001, hanMarkovChainMonte2001}, allowing for quantification of uncertainty, individual-level probability estimates, and the integration of additional data sources \citep{mccormickProbabilisticCauseofDeathAssignment2016b}.GPT-3.5 \citep{brownLanguageModelsAre2020} and GPT-4 \citep{openaiGPT4TechnicalReport2023} are large language models (LLMs) based on transformer architectures \citep{vaswaniAttentionAllYou2017}. These models are trained using reinforcement learning from human feedback \citep{ouyangTrainingLanguageModels2022, christianoDeepReinforcementLearning2017, stiennonLearningSummarizeHuman2020, wirthSurveyPreferencebasedReinforcement2017}, enabling them to follow natural language instructions and generate human-level responses. GPT-4 introduces improvements over GPT-3.5, including more recent training data, enhanced reasoning capabilities, and multimodal input-output functionality (e.g. text, image, voice) \citep{wuBriefOverviewChatGPT2023}.

>>>>

VA involves two major components: survey and COD assignment \citep{worldhealthorganizationVerbalAutopsyStandards2023,chandramohanEstimatingCausesDeath2021,worldhealthorganizationVerbalAutopsyStandards2007}. In the survey component, trained interviewers use structured questionnaires and open narrative prompts to gather data from relatives or close contacts of the deceased. In the COD assignment component, physicians review these data to determine the most likely COD. However, reliance on physician assignment has been criticized for limited reproducibility and subjectivity \citep{gomesNationwideMortalityStudies2017,jhaProspectiveStudyOne2005,mccormickProbabilisticCauseofdeathAssignment2016b,morrisFactorsAssociatedPhysician2010,solemanVerbalAutopsyCurrent2006}. To overcome these limitations, automated Computer Coded Verbal Autopsy (CCVA) methods such as InterVA \citep{byassIntegratedApproachProcessing2019} and InSilicoVA \citep{mccormickProbabilisticCauseofdeathAssignment2016b} have been developed. These models offer scalable and reproducible alternatives and have demonstrated comparable performance to physicians at the population level. However, their performance at the individual level remains limited \citep{jhaAutomatedPhysicianAssignment2019,leitaoComparisonPhysiciancertifiedVerbal2014,desaiPerformanceFourComputercoded2014,tungaVerbalAutopsyModels2021,otiVerbalAutopsyInterpretation2010}, while their reliance on structured questionnaire data often omits open narrative text, which can contain additional contextual and chronological information that may improve diagnostic accuracy \citep{jebleeAutomaticallyDeterminingCause2019,blancoExtractingCauseDeath2021,kingQualityDiagnosticValue2016}.

@@

Four computational models were used to assign causes of death (CODs) for each of the 6,939 physician-agreed verbal autopsy (VA) records: GPT-3.5, GPT-4, InterVA-5, and InSilicoVA. InterVA-5 and InSilicoVA are widely used statistical models within the OpenVA framework for COD assignment in VAs \citep{murrayUsingVerbalAutopsy2014, otiVerbalAutopsyInterpretation2010, benaraEvaluationMethodsAssigning2023, chandramohanEstimatingCausesDeath2021, tungaVerbalAutopsyModels2021, leitaoComparisonPhysiciancertifiedVerbal2014, jhaAutomatedPhysicianAssignment2019, liOpenVAToolkitVerbal2023}. InterVA-5 applies a Bayesian probabilistic approach, using a standardized set of symptoms and expert-derived conditional probabilities to assign the most likely COD based on maximum probability \citep{byassIntegratedApproachProcessing2019, byassStrengtheningStandardisedInterpretation2012, bayesEssaySolvingProblem1958}. InSilicoVA extends this approach by incorporating a hierarchical Bayesian framework and Markov Chain Monte Carlo (MCMC) methods \citep{brooksMarkovChainMonte1998, chibMarkovChainMonte2001, hanMarkovChainMonte2001}, allowing for quantification of uncertainty, individual-level probability estimates, and the integration of additional data sources \citep{mccormickProbabilisticCauseofdeathAssignment2016b}.GPT-3.5 \citep{brownLanguageModelsAre2020} and GPT-4 \citep{openaiGPT4TechnicalReport2023} are large language models (LLMs) based on transformer architectures \citep{vaswaniAttentionAllYou2017}. These models are trained using reinforcement learning from human feedback \citep{ouyangTrainingLanguageModels2022, christianoDeepReinforcementLearning2017, stiennonLearningSummarizeHuman2020, wirthSurveyPreferencebasedReinforcement2017}, enabling them to follow natural language instructions and generate human-level responses. GPT-4 introduces improvements over GPT-3.5, including more recent training data, enhanced reasoning capabilities, and multimodal input-output functionality (e.g. text, image, voice) \citep{wuBriefOverviewChatGPT2023}.
